---
title: "FTLD Interactive Results"
author: "Joaquim Aumatell"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

```{r libraries}
library(dplyr)
library(tidyr)
library(readr)
library(plotly)
library(purrr)
library(stringr)
library(glue)
library(tibble)
library(htmltools)
```

```{r helpers}
set.seed(123)

allowed_extensions <- c("csv", "tsv", "txt", "rds", "rda", "rdata", "xlsx", "xls")

resolve_result_path <- function(option_name, pattern = NULL) {
  opt_path <- getOption(option_name)
  if (!is.null(opt_path) && file.exists(opt_path)) {
    return(normalizePath(opt_path))
  }
  env_name <- toupper(gsub("\\.", "_", option_name))
  env_path <- Sys.getenv(env_name, unset = "")
  if (nzchar(env_path) && file.exists(env_path)) {
    return(normalizePath(env_path))
  }
  if (!is.null(pattern)) {
    matches <- list.files(path = Sys.getenv("RESULTS_VIS_ROOT", unset = "."),
                          pattern = pattern,
                          full.names = TRUE,
                          recursive = TRUE,
                          ignore.case = TRUE)
    matches <- matches[tolower(tools::file_ext(matches)) %in% allowed_extensions]
    if (length(matches) > 0) {
      return(normalizePath(matches[1]))
    }
  }
  NULL
}

read_result_file <- function(path) {
  if (is.null(path)) {
    return(NULL)
  }
  ext <- tolower(tools::file_ext(path))
  tryCatch({
    if (ext == "csv") {
      readr::read_csv(path, show_col_types = FALSE)
    } else if (ext %in% c("tsv", "txt")) {
      readr::read_tsv(path, show_col_types = FALSE)
    } else if (ext == "rds") {
      readRDS(path)
    } else if (ext %in% c("rda", "rdata")) {
      env <- new.env(parent = emptyenv())
      objs <- load(path, envir = env)
      if (length(objs) == 0) {
        NULL
      } else {
        env[[objs[1]]]
      }
    } else if (ext %in% c("xlsx", "xls")) {
      if (!requireNamespace("readxl", quietly = TRUE)) {
        warning("Package 'readxl' is required to import Excel files. Returning NULL so that placeholder data are used instead.")
        return(NULL)
      }
      readxl::read_excel(path)
    } else {
      stop("Unsupported file extension: ", ext)
    }
  }, error = function(e) {
    warning("Failed to read data from ", path, ": ", conditionMessage(e))
    NULL
  })
}

load_or_placeholder <- function(option_name, pattern, generator) {
  path <- resolve_result_path(option_name, pattern)
  data <- read_result_file(path)
  if (!is.null(data)) {
    attr(data, "placeholder") <- FALSE
    attr(data, "source") <- path
    return(data)
  }
  example <- generator()
  attr(example, "placeholder") <- TRUE
  attr(example, "source") <- "placeholder"
  example
}

render_source_note <- function(data, description) {
  placeholder <- isTRUE(attr(data, "placeholder"))
  source_path <- attr(data, "source")
  if (placeholder) {
    htmltools::div(class = "placeholder-note",
                   htmltools::tags$em(glue::glue("No external {description} file detected; displaying placeholder data.")))
  } else {
    htmltools::div(class = "source-note",
                   htmltools::tags$strong("Source: "),
                   htmltools::tags$code(source_path))
  }
}

annotate_placeholder <- function(plot_obj, data) {
  if (!isTRUE(attr(data, "placeholder"))) {
    return(plot_obj)
  }
  plot_obj %>% layout(annotations = list(
    text = "Placeholder example — update the data source to view real results",
    x = 0.5,
    y = 1.05,
    xref = "paper",
    yref = "paper",
    showarrow = FALSE,
    font = list(color = "#cc0000", size = 12)
  ))
}
```

The next sections follow the same numbering as the comprehensive workflow document.

## 0. SVA

Estimate surrogate variables (SVs) for each cohort and visualise their distribution. The expected input is a table with at least `SV1`, `SV2`, and cohort annotations.

```{r sva-data}
sva_results <- load_or_placeholder(
  option_name = "results_vis.sva",
  pattern = "sva.*(scores|results).*\\.(csv|tsv|txt|rds|rda|RData|xlsx)$",
  generator = function() {
    tibble::tibble(
      Sample = paste0("Sample", seq_len(20)),
      Cohort = rep(c("C9", "Pottier", "Rimmod", "TDP", "Tau"), each = 4),
      SV1 = rnorm(20),
      SV2 = rnorm(20)
    )
  }
)
```

```{r sva-plot}
render_source_note(sva_results, "SVA")
plot_ly(
  sva_results,
  x = ~SV1,
  y = ~SV2,
  color = ~Cohort,
  text = ~Sample,
  type = "scatter",
  mode = "markers"
) %>%
  layout(title = "Surrogate Variable Structure",
         xaxis = list(title = "SV1"),
         yaxis = list(title = "SV2")) %>%
  annotate_placeholder(sva_results)
```

## 1. Bulk DEA

Interactive volcano plot summarising differential expression results from bulk RNA-seq analyses.

```{r bulk-dea-data}
bulk_dea <- load_or_placeholder(
  option_name = "results_vis.bulk_dea",
  pattern = "bulk.*dea.*\\.(csv|tsv|txt|rds|rda|RData|xlsx)$",
  generator = function() {
    tibble::tibble(
      Gene = paste0("Gene", seq_len(300)),
      logFC = rnorm(300, sd = 1.2),
      pvalue = runif(300, min = 1e-6, max = 0.5),
      Comparison = sample(c("C9 vs Control", "TDP vs Control", "Tau vs Control"), 300, replace = TRUE)
    ) %>%
      mutate(adj.P.Val = p.adjust(pvalue, method = "BH"),
             negLogP = -log10(pvalue))
  }
)
```

```{r bulk-dea-plot}
render_source_note(bulk_dea, "bulk DEA")
plot_ly(
  bulk_dea,
  x = ~logFC,
  y = ~negLogP,
  color = ~Comparison,
  text = ~paste("Gene:", Gene, "<br>adj.P.Val:", signif(adj.P.Val, 3)),
  type = "scatter",
  mode = "markers",
  hovertemplate = "%{text}<extra></extra>"
) %>%
  layout(title = "Bulk Differential Expression Volcano",
         xaxis = list(title = "log2 Fold Change"),
         yaxis = list(title = "-log10(p-value)")) %>%
  annotate_placeholder(bulk_dea)
```

## 2. BayesPrism

View cell-type composition estimates from the BayesPrism deconvolution output.

```{r bayesprism-data}
bayesprism <- load_or_placeholder(
  option_name = "results_vis.bayesprism",
  pattern = "bayesprism.*(composition|proportion).*\\.(csv|tsv|txt|rds|rda|RData|xlsx)$",
  generator = function() {
    tibble::tibble(
      Sample = rep(paste0("Sample", seq_len(12)), each = 5),
      CellType = rep(c("Astrocyte", "Neuron", "Microglia", "Oligodendrocyte", "Endothelial"), times = 12),
      Proportion = runif(60)
    ) %>%
      group_by(Sample) %>%
      mutate(Proportion = Proportion / sum(Proportion)) %>%
      ungroup()
  }
)
```

```{r bayesprism-plot}
render_source_note(bayesprism, "BayesPrism")
plot_ly(
  bayesprism,
  x = ~Sample,
  y = ~Proportion,
  color = ~CellType,
  type = "bar"
) %>%
  layout(barmode = "stack",
         title = "Cell-Type Proportions per Sample",
         xaxis = list(title = "Sample"),
         yaxis = list(title = "Proportion", range = c(0, 1))) %>%
  annotate_placeholder(bayesprism)
```

## 3. Cell Proportion Differences

Compare estimated cell-type proportions between diagnostic groups.

```{r cell-prop-diff-data}
cell_prop_diff <- load_or_placeholder(
  option_name = "results_vis.cell_prop_diff",
  pattern = "cell.*prop.*(diff|difference).*\\.(csv|tsv|txt|rds|rda|RData|xlsx)$",
  generator = function() {
    tibble::tibble(
      Group = rep(c("Control", "FTLD"), each = 50),
      CellType = rep(c("Astrocyte", "Neuron", "Microglia", "Oligodendrocyte", "Endothelial"), each = 20),
      Proportion = rbeta(100, shape1 = 2, shape2 = 5)
    )
  }
)
```

```{r cell-prop-diff-plot}
render_source_note(cell_prop_diff, "cell proportion differences")
plot_ly(
  cell_prop_diff,
  x = ~Group,
  y = ~Proportion,
  color = ~CellType,
  type = "box"
) %>%
  layout(title = "Cell Proportion Differences by Group",
         yaxis = list(title = "Estimated Proportion")) %>%
  annotate_placeholder(cell_prop_diff)
```

## 4. Cell Proportion Correlations

Heatmap of correlations between cell-type proportions and relevant phenotypes or modules.

```{r cell-prop-corr-data}
cell_prop_corr <- load_or_placeholder(
  option_name = "results_vis.cell_prop_corr",
  pattern = "cell.*prop.*(corr|correlation).*\\.(csv|tsv|txt|rds|rda|RData|xlsx)$",
  generator = function() {
    cell_types <- c("Astrocyte", "Neuron", "Microglia", "Oligodendrocyte", "Endothelial")
    metrics <- c("Age", "DiseaseDuration", "MMSE", "Module1", "Module2")
    expand_grid(CellType = cell_types, Metric = metrics) %>%
      mutate(Correlation = runif(n(), min = -1, max = 1))
  }
)
```

```{r cell-prop-corr-plot}
render_source_note(cell_prop_corr, "cell proportion correlations")
cell_prop_corr_wide <- cell_prop_corr %>%
  tidyr::pivot_wider(names_from = Metric, values_from = Correlation)
cell_prop_corr_matrix <- cell_prop_corr_wide %>%
  column_to_rownames("CellType") %>%
  as.matrix()
cell_prop_corr_hover <- cell_prop_corr %>%
  mutate(label = glue::glue("{CellType} vs {Metric}<br>Correlation: {round(Correlation, 2)}")) %>%
  tidyr::pivot_wider(names_from = Metric, values_from = label) %>%
  column_to_rownames("CellType") %>%
  as.matrix()
plot_ly(
  x = colnames(cell_prop_corr_matrix),
  y = rownames(cell_prop_corr_matrix),
  z = cell_prop_corr_matrix,
  type = "heatmap",
  colors = "RdBu",
  reversescale = TRUE,
  text = cell_prop_corr_hover,
  hoverinfo = "text"
) %>%
  layout(title = "Correlation of Cell Proportions with Metadata") %>%
  annotate_placeholder(cell_prop_corr)
```

## 5. Cell Type DEA

Interactive summary of differential expression performed on single-cell clusters or pseudobulk aggregates.

```{r cell-type-dea-data}
cell_type_dea <- load_or_placeholder(
  option_name = "results_vis.cell_type_dea",
  pattern = "cell.*type.*dea.*\\.(csv|tsv|txt|rds|rda|RData|xlsx)$",
  generator = function() {
    tibble::tibble(
      Cluster = rep(paste("Cluster", 1:8), each = 40),
      Gene = paste0("Gene", seq_len(320)),
      avg_log2FC = rnorm(320, sd = 0.8),
      pct_in = runif(320, min = 0, max = 1),
      pct_out = runif(320, min = 0, max = 1),
      p_val_adj = p.adjust(runif(320, min = 1e-5, max = 0.5), method = "BH")
    )
  }
)
```

```{r cell-type-dea-plot}
render_source_note(cell_type_dea, "cell type DEA")
plot_ly(
  cell_type_dea,
  x = ~avg_log2FC,
  y = ~-log10(p_val_adj + 1e-12),
  color = ~Cluster,
  text = ~paste(
    "Gene:", Gene,
    "<br>pct_in:", sprintf("%.1f%%", pct_in * 100),
    "<br>pct_out:", sprintf("%.1f%%", pct_out * 100)
  ),
  type = "scatter",
  mode = "markers",
  hovertemplate = "%{text}<extra></extra>"
) %>%
  layout(title = "Cell-Type Differential Expression",
         xaxis = list(title = "Average log2 Fold Change"),
         yaxis = list(title = "-log10(adj p-value)")) %>%
  annotate_placeholder(cell_type_dea)
```

## 6. Genes of Interest Correlation

Inspect correlations among curated gene sets or between genes and phenotypic scores.

```{r goi-corr-data}
goi_corr <- load_or_placeholder(
  option_name = "results_vis.goi_corr",
  pattern = "genes.*(interest|goi).*corr.*\\.(csv|tsv|txt|rds|rda|RData|xlsx)$",
  generator = function() {
    genes <- paste0("Gene", LETTERS[1:10])
    metrics <- paste0("Metric", 1:10)
    expand_grid(Gene = genes, Metric = metrics) %>%
      mutate(Correlation = runif(n(), min = -1, max = 1))
  }
)
```

```{r goi-corr-plot}
render_source_note(goi_corr, "genes of interest correlation")
goi_wide <- goi_corr %>%
  tidyr::pivot_wider(names_from = Metric, values_from = Correlation)
goi_matrix <- goi_wide %>%
  column_to_rownames("Gene") %>%
  as.matrix()
goi_hover <- goi_corr %>%
  mutate(label = glue::glue("{Gene} vs {Metric}<br>Correlation: {round(Correlation, 2)}")) %>%
  tidyr::pivot_wider(names_from = Metric, values_from = label) %>%
  column_to_rownames("Gene") %>%
  as.matrix()
plot_ly(
  x = colnames(goi_matrix),
  y = rownames(goi_matrix),
  z = goi_matrix,
  type = "heatmap",
  colors = "RdBu",
  reversescale = TRUE,
  text = goi_hover,
  hoverinfo = "text"
) %>%
  layout(title = "Genes of Interest Correlation Matrix") %>%
  annotate_placeholder(goi_corr)
```

## 7. HDWGCNA

Visualise module eigengene relationships from the HDWGCNA analysis.

```{r hdwgcna-data}
hdwgcna <- load_or_placeholder(
  option_name = "results_vis.hdwgcna",
  pattern = "hdwgcna.*(module|eigengene|trait).*\\.(csv|tsv|txt|rds|rda|RData|xlsx)$",
  generator = function() {
    modules <- paste0("Module", seq_len(8))
    traits <- c("Diagnosis", "Age", "Sex", "Region", "PMI")
    expand_grid(Module = modules, Trait = traits) %>%
      mutate(Correlation = runif(n(), min = -1, max = 1),
             Pvalue = runif(n(), min = 1e-4, max = 0.2))
  }
)
```

```{r hdwgcna-plot}
render_source_note(hdwgcna, "HDWGCNA")
hdwgcna_wide <- hdwgcna %>%
  tidyr::pivot_wider(names_from = Trait, values_from = Correlation)
hdwgcna_matrix <- hdwgcna_wide %>%
  column_to_rownames("Module") %>%
  as.matrix()
hdwgcna_hover <- hdwgcna %>%
  mutate(label = glue::glue("{Module} vs {Trait}<br>Correlation: {round(Correlation, 2)}<br>p-value: {signif(Pvalue, 2)}")) %>%
  tidyr::pivot_wider(names_from = Trait, values_from = label) %>%
  column_to_rownames("Module") %>%
  as.matrix()
plot_ly(
  x = colnames(hdwgcna_matrix),
  y = rownames(hdwgcna_matrix),
  z = hdwgcna_matrix,
  type = "heatmap",
  colors = "RdBu",
  reversescale = TRUE,
  text = hdwgcna_hover,
  hoverinfo = "text"
) %>%
  layout(title = "HDWGCNA Module-Trait Relationships") %>%
  annotate_placeholder(hdwgcna)
```

## 8. Differences in Modules

Summaries of module activity differences between diagnostic groups.

```{r module-diff-data}
module_diff <- load_or_placeholder(
  option_name = "results_vis.module_diff",
  pattern = "module.*(difference|activity).*\\.(csv|tsv|txt|rds|rda|RData|xlsx)$",
  generator = function() {
    tibble::tibble(
      Module = paste0("Module", seq_len(12)),
      Group = rep(c("Control", "FTLD"), each = 12),
      Score = rnorm(24, mean = rep(c(0, 1), each = 12), sd = 0.5)
    )
  }
)
```

```{r module-diff-plot}
render_source_note(module_diff, "module differences")
plot_ly(
  module_diff,
  x = ~Module,
  y = ~Score,
  color = ~Group,
  type = "bar"
) %>%
  layout(barmode = "group",
         title = "Module Activity Differences",
         xaxis = list(title = "Module"),
         yaxis = list(title = "Score")) %>%
  annotate_placeholder(module_diff)
```

## 9. Correlation Modules

Interactive network-style heatmap of module-module correlations.

```{r correlation-modules-data}
correlation_modules <- load_or_placeholder(
  option_name = "results_vis.correlation_modules",
  pattern = "module.*correlation.*\\.(csv|tsv|txt|rds|rda|RData|xlsx)$",
  generator = function() {
    modules <- paste0("Module", seq_len(10))
    expand_grid(Source = modules, Target = modules) %>%
      mutate(Correlation = if_else(Source == Target, 1, runif(n(), min = -1, max = 1)))
  }
)
```

```{r correlation-modules-plot}
render_source_note(correlation_modules, "module correlations")
correlation_wide <- correlation_modules %>%
  tidyr::pivot_wider(names_from = Target, values_from = Correlation)
correlation_matrix <- correlation_wide %>%
  column_to_rownames("Source") %>%
  as.matrix()
correlation_hover <- correlation_modules %>%
  mutate(label = glue::glue("{Source} vs {Target}<br>Correlation: {round(Correlation, 2)}")) %>%
  tidyr::pivot_wider(names_from = Target, values_from = label) %>%
  column_to_rownames("Source") %>%
  as.matrix()
plot_ly(
  x = colnames(correlation_matrix),
  y = rownames(correlation_matrix),
  z = correlation_matrix,
  type = "heatmap",
  colors = "RdBu",
  reversescale = TRUE,
  text = correlation_hover,
  hoverinfo = "text"
) %>%
  layout(title = "Module-Module Correlation Matrix") %>%
  annotate_placeholder(correlation_modules)
```

## 10. Validation

Assess concordance metrics or overlap statistics between datasets in an interactive format.

```{r validation-data}
validation_results <- load_or_placeholder(
  option_name = "results_vis.validation",
  pattern = "validation.*(summary|metrics|overlap).*\\.(csv|tsv|txt|rds|rda|RData|xlsx)$",
  generator = function() {
    tibble::tibble(
      Comparison = rep(c("C9 vs External", "TDP vs External", "Tau vs External"), each = 15),
      Metric = rep(c("Jaccard", "Overlap", "AUROC"), times = 15),
      Value = runif(45, min = 0.2, max = 0.95)
    )
  }
)
```

```{r validation-plot}
render_source_note(validation_results, "validation")
plot_ly(
  validation_results,
  x = ~Metric,
  y = ~Value,
  color = ~Comparison,
  type = "box"
) %>%
  layout(title = "Validation Metrics Across Comparisons",
         yaxis = list(title = "Metric Value")) %>%
  annotate_placeholder(validation_results)
```

## Session Information

```{r session-info}
sessionInfo()
```

